\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}

### 2013-12-03 docs/paper/paper.tex
bezig met het stukje over causal faithfulness i.c.m. PC

We define a trail as a set of edges which would make up a path if those edges would be undirected.
We can now define directional separation, or d-separation.
Two nodes $X$ and $Y$ are d-separated given a set of nodes $S$ (with $X, Y \notin S$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
\begin{itemize}
\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $S$.
\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $S$.
\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $S$.
\end{itemize}
If $X$ and $Y$ are d-separated given $S$, $S$ is called a separating set of $X$ and $Y$. Every two independent variables have a separating set; if they are unconditionally independent, his is the empty set.

\subsection{The PC-algorithm}
$\srightarrow$
$\sleftarrow$
$\srightline$
$\sleftline$
$\sline$
\cite{?}
Whether there is a dependency between two variables can be concluded by using a statistical test.
In this way, a graph can be inferred representing all dependencies between nodes.
However, this only results in an undirected graph, as it does not give any information about causality.
A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.

One of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent, .
As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
If $Z$ is not in $S$, then the only possible causal relation can be that of a v-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.

The other rules are based on arrows that have already been found.
\begin{itemize}
\item If there is an edge $X \sline Y$, and if there is a directed path from $X$ to $Y$, then $X \sline Y$ can be oriented as $X \srightarrow Y$.
\item If there are edges $X \srightarrow Y$ and $Y \sleftline Z$, if the latter is not an edge $Y \sleftarrow Z$, and if $X$ and $Z$ are not connected, orient $Y \sline Z$ as $Y \rightarrow Z$. 
\end{itemize}

>>>>

We define a trail as a set of edges which would make up a path if those edges would be undirected.
We can now define directional separation, or d-separation.
Two nodes $X$ and $Y$ are d-separated given a set of nodes $S$ (with $X, Y \notin S$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
\begin{itemize}
\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $S$.
\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $S$.
\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $S$.
\end{itemize}
If $X$ and $Y$ are d-separated given $S$, $S$ is called a separating set of $X$ and $Y$. Every two independent variables have a separating set; if they are unconditionally independent, this is the empty  set.

\subsection{Assumptions about the structure}
In theory, every deterministic real-world situation must be described by some Bayesian model, as all situations conform to causality.
But the underlying real-world data such as the measured brain activity used in this analysis usually does not conform to a simple model.
Assumption can be made about the model to make it easier to infer it, but this increases the risk of drawing false conclusions.

One assumption that can be made about the model is that of causal sufficiency. 
When it is assumed that the model fully describes the real network, so-called V-structures can be concluded. 
When there is a structure $X - Y - Z$, in which $X$ and $Z$ are not connected, the only possible orientation of the connections is $X \rightarrow Y \leftarrow Z$. % TODO: extra uitleg?
An approach to finding a model is therefore to find all V-structures and orient them in this way.

However, it is very unlikely that all events which are relevant for the causal graph are measured; 
even with high temporal and spatial resolutions, the entire fysiology of the brain is too complex to capture.
Latent variables are nodes which are present in the underlying Bayesian graph, but which are not measured.
When there is a structure $X \leftarrow L \rightarrow Y$ in which $L$ is a latent variable, an edge between $X$ and $Y$ might be found while the these two nodes have no direct ancestor/descendant-relation between them.
A directed edge $X \rightarrow Y$ or $X \leftarrow Y$ would be erroneous, so an arrow 
The complex reality might be very similar to a simple model based on the measured data, but it might also be very different.
This correctness of the data is called causal faithfulness.
As such, the way of describing V-structures might not draw false conclusions, depending on the causal faithfulness.
When causal sufficiency is not assumed,

\subsection{The PC-algorithm}
\cite{?}
Whether there is a dependency between two variables can be concluded by using a statistical test.
In this way, a graph can be inferred representing all dependencies between nodes.
However, this only results in an undirected graph, as it does not give any information about causality.
A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm. Multiple variations of this algorithm are described by Spirtes, Glymour and Scheines \ref{?}. For this analysis, we have used the "modified pc algorithm" which does not assume causal sufficiency, and thus is able to handle latent variables.

The first of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent.
As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
If $Z$ is not in $S$, then the only possible causal relation can be that of a v-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.

After this rules which finds V-structures, the other rules are applied in no particular order. These rules are based on arrows that have already been found.
\begin{itemize}
\item If there is an edge $X \sline Y$, and if there is a directed path from $X$ to $Y$, then $X \sline Y$ can be oriented as $X \srightarrow Y$.
\item If there are edges $X \srightarrow Y$ and $Y \sleftline Z$, if the latter is not an edge $Y \sleftarrow Z$, and if $X$ and $Z$ are not connected, orient $Y \sline Z$ as $Y \rightarrow Z$. 
\end{itemize}
